---
title: "Activity monitoring data report"
author: "Gabriel Sebastião"
date: "16/01/2020"
output: html_document
---

## Loading and preprocessing the data

First the data is loaded, and if needed the zip file is unziped:
```{r Data Loading}
if (!file.exists("activity.csv")){
        unzip("activity.zip") 
}
act_data <- read.csv("activity.csv",na.strings = "NA")
act_data$date <- as.Date(act_data$date, format = "%Y-%m-%d")

```

## What is mean total number of steps taken per day?

Calculating the total number of steps taken per day:

```{r total number of steps}
total_step_day <- aggregate(steps ~ date,act_data, sum)
head(total_step_day)
```

Then, the histogram is drawn:

```{r Histogram, fig.height = 4}
hist(total_step_day$steps, main = "Histogram of total steps taken in a day", xlab = "Total Steps")
```

## What is the average daily activity pattern?

The mean and median values of total number of steps taken per day are:

```{r Mean and median}
step_mean <- mean(total_step_day$steps)
step_median <- median(total_step_day$steps)
# The mean is:
step_mean
# And the median:
step_median
```

```{r Time series graph, fig.height = 4}
total_step_interval <- aggregate(steps ~ interval,act_data, mean)
plot(total_step_interval, type = "l", main = "Time series graph")
```

The interval with the maximum number of steps is:  
```{r Max interval}
total_step_interval$interval[max(total_step_interval$steps)]
```
Therefore, the interval of 5:05pm is the day with the maximum number of steps, on average.

# Imputing missing values

Calculating how much of the data is missing:

```{r Calculating how much of the data is missing}
sum(is.na(act_data))
(mean(is.na(act_data))*100)
```

We arrive at the percentual of missing data: 4.37%

### Dealing with the missing values

Since it was suggested the creation of a new dataset, the approach chosen was to input the average value of steps for the interval, when there's a missing value

```{r creating the new dataset}
data_mean_interval <- act_data
for (i in 1:nrow(data_mean_interval)) {
  if (is.na(data_mean_interval$steps[i])) {
    interval_match <- data_mean_interval$interval[i]
    row_match <- total_step_interval[total_step_interval$interval == interval_match,]
    data_mean_interval$steps[i] <- row_match$steps
  }
}
```

Now we can calculate the total number of steps taken per day and see the results:

```{r}
total_step_day_filled <- aggregate(steps ~ date, data_mean_interval, sum)
head(total_step_day_filled)
```

Now there are values for January 1st for example, let's see if the histogram has changed as well:

```{r NewHistogram, fig.height = 4}
hist(total_step_day_filled$steps, main = "Histogram of total steps taken in a day", xlab = "Total Steps")
```

Well, it doesn't seem any different, let's look at the mean and median values:

```{r 2nd Mean and median}
step_mean_filled <- mean(total_step_day_filled$steps)
step_median_filled <- median(total_step_day_filled$steps)
# The mean is:
step_mean_filled
# And the median:
step_median_filled
```
Well, both the mean and the median have converged into the value of the first mean that was taken before dealing with the NAs. So, in a way the dataset was filled in such way that there wasn't any divergence towards the mean values.

# Are there differences in activity patterns between weekdays and weekends?


```{r 2nd Time series graph, fig.height = 4}

library(lubridate)
library(ggplot2)
for (i in 1:nrow(data_mean_interval)) {
        if (wday(data_mean_interval$date[i]) == 1 | wday(data_mean_interval$date[i]) == 7){
                data_mean_interval$day_type[i] <- "weekend"
        }
        else{
                data_mean_interval$day_type[i] <- "weekday"
        }
}

data_mean_interval$day_type <- as.factor(data_mean_interval$day_type)

total_step_interval_filled <- aggregate(steps ~ interval + day_type, data_mean_interval, mean)

qplot(interval, steps ,data = total_step_interval_filled, facets = day_type~., type = "l", geom=c("line"), main = "Time series graph")
```
